---
title: "Untitled"
format: html
editor: visual
---

```{r}
library('tidyverse')
```

------------------------------------------------------------------------

Changed 171-1 to 171 on row 1522 column 1 in "Transplanting_exp.csv". Removed row 1348-1353.

```{r}
tplant <- read_csv(file = "Transplanting_exp.csv", 
                   col_types = cols(
                     ID = col_character(),
                     Pop = col_integer(),
                     Replicate = col_character(),
                     Tray = col_factor(),
                     Row = col_integer(),
                     Column = col_integer(),
                     Harvest_date = col_character(),
                     FT = col_character()
                   ))

tplant
```

```{r}
# Summing up replicats for each population
summed_pop <- tplant |>
  group_by(Pop) |>
  summarise(total_repl = n()) 

summed_pop

```

```{r}
# add label for threshold of total replicats above 5
summed_pop_above_4 <- summed_pop |>
  mutate(pop_above_4 = ifelse(total_repl < 4, FALSE, TRUE))

summed_pop_above_4
```

```{r}
summed_pop_above_4 |>
  filter(pop_above_4 == T) |>
  nrow()
```

Out of the 162 populations 114 have more than 4 replicates, 48 less.

The following Population have not 5 replicates.

```{r}
# Populations below threshold of replicates
tplant_pop_below_4 <- summed_pop_above_4 |>
  filter(pop_above_4 == F) 

tplant_pop_below_4$Pop
```

------------------------------------------------------------------------

In the data set **Flowering_time_in-situ** you have the FT of the population IN THE CITY - in-situ.

You are interested in the column:

\- **Site_ID** the population number

\- **Type** the type of urban habitat

\- **First_FT_julianday**
the first flowering individual of the population transformed into
Julian day (counting the day of the year from 1 to 365, 1 being the 1st
of January and 365 the 31th December). It is the F3 generation here.

```{r}
ft_situ <- read_csv(file = "Flowering_time_in-situ.csv")

ft_situ
```

------------------------------------------------------------------------

In the data set **FT_table_final** it
is Gaëlle data set of the FT CONTROLLED CONDITIONS - ex-situ. She was
working with the generation F4 sampled in 2023 and **you are working with
the generation F5** from 2025

\- **Column A - H** are the same as your data set!

To extract the FT of the date set you need to do **Sowing date - Flowering date**

Gaëlle also compare the FT in-situ and ex-situ this is the F3/F4 correlation at the end of her script.

```{r}
f4_gaelle <- read_csv(file = "FT_table_final.csv")

f4_gaelle
```

```{r}
# A-H are same so reproducing ID from tplant to join
colnames(f4_gaelle) <- paste0(colnames(f4_gaelle), "_F4")
f4_gaelle_w_id <- f4_gaelle |>
  mutate(ID = paste0(as.character(Population_F4), "-", as.character(Replicate_F4)))



tplant_w_f4 <- tplant |>
  left_join(f4_gaelle_w_id, by = "ID")

tplant_w_f4
```
