---
title: "Untitled"
format: html
editor: visual
---

```{r}
#| message: false
library('tidyverse')
```

------------------------------------------------------------------------

```{r}
tplant <- read_csv(file = "data/tplant_observations - main.csv",
                   na = "Na",
                   col_types = cols(
                     ID = col_character(),
                     Tray = col_factor(),
                     Row = col_integer(),
                     Column = col_integer(),
                     n_shoot_axis = col_integer(),
                     Leaf_shape = col_factor(),
                     Leaf_edge = col_factor(),
                     Absent = col_logical(),
                     is_A_tha = col_logical(),
                     Comments = col_character()
                   )) |>
  filter(!(is.na(ID))) |>
  mutate(Pop = stringi::stri_extract(ID, regex = "^.*(?=-)"),
         Replicate = stringi::stri_extract(ID, regex = "(?<=-).*$")) |>
  select(ID, Pop, Replicate, everything()) 

tplant
```

```{r}
# Summing up replicats for each population
summed_pop <- tplant |>
  filter(is_A_tha %in% c(T, NA),
         Absent %in% c(F, NA))|>
  group_by(Pop) |>
  summarise(total_repl = n()) |>
  arrange(desc(total_repl))

summed_pop
```

```{r}
# add label for threshold of total replicats above 5
summed_pop_above_4 <- summed_pop |>
  mutate(pop_above_4 = ifelse(total_repl > 4, T, F))

summed_pop_above_4
```

```{r}
summed_pop_above_4 |>
  filter(pop_above_4 == T) |>
  nrow()
```

Out of the 164 populations 95 have more than 4 replicates, x less.

The following Population have not 5 replicates.

```{r}
# Populations below threshold of replicates
tplant_pop_below_4 <- summed_pop_above_4 |>
  filter(pop_above_4 == F) 

tplant_pop_below_4$Pop
```

```{r}
# Add the above to the tplant df
tplant <- tplant |>
  group_by(Pop) |>
  filter(is_A_tha %in% c(T, NA),
         Absent %in% c(F, NA))|>
  mutate(total_repl_in_pop = n()) |>
  ungroup() |>
  mutate(pop_above_4 = ifelse(total_repl_in_pop > 4, T, F)) 


tplant
```

What Populations are in the first Batch, primarily Tray 1-7:

```{r}
Pop_of_batch_1 <- tplant |>
  filter(Tray == 1:7) |>
  select(Pop) 


tplant_batch_1 <- tplant |>
  filter(Pop %in% Pop_of_batch_1$Pop) |>
  arrange(Pop) |>
  filter(pop_above_4 == T)

tplant_batch_1
```

------------------------------------------------------------------------

In the data set **Flowering_time_in-situ** you have the FT of the population IN THE CITY - in-situ.

You are interested in the column:

\- **Site_ID** the population number

\- **Type** the type of urban habitat

\- **First_FT_julianday** the first flowering individual of the population transformed into Julian day (counting the day of the year from 1 to 365, 1 being the 1st of January and 365 the 31th December). It is the F3 generation here. The **first Julianday for 01.01.2023 is 2459946**.

```{r}
ft_insitu_raw <- read_csv(file = "data/Flowering_time_in-situ.csv",
                    col_types = cols(
                      Site_ID = col_character(),
                      Type = col_factor()
                      )
                    )

missing_type <- read_csv(file = "data/Missing_type.csv",
                         col_types = cols(
                           Site_ID = col_character(),
                           Type = col_factor()
                         ))

First_FT_julianday_raw <- read_csv(file = "data/Phenology_field.csv",
                               col_types = cols(
                                 Site_ID = col_character(),
                                 Type = col_factor(),
                                 First_FT_julianday = col_double()
                               )) |>
  select(Site_ID, Type, First_FT_julianday)

# join the raw in-situ table with the missing types. Add the total days from julianday to First_FT_Julianday
ft_insitu_with_pheno_raw <- ft_insitu_raw |>
  rows_append(missing_type)|>
  left_join(First_FT_julianday_raw, by = "Site_ID", suffix = c("_insitu", "_Pheno_field"))

# add First_FT_julianday from Phenology_field df where missing in FT_insitu
ft_insitu <- ft_insitu_with_pheno_raw |>
  mutate(
    First_FT_julianday_raw = ifelse(
      is.na(First_FT_julianday_insitu),
      First_FT_julianday_Pheno_field,
      First_FT_julianday_insitu
    )
  ) |> 
  mutate(First_FT_julianday = First_FT_julianday_raw + 2459946,
         Type = Type_insitu) |>
  select(-Type_Pheno_field, -First_FT_julianday_Pheno_field, -First_FT_julianday_insitu, -Type_insitu)

ft_insitu
```

Join the in-situ data with the transplanted plants (tplant) data set by population where the column `Site_ID` is equal to the column `Population`.

```{r}
pop_per_env <- summed_pop_above_4 |>
  left_join(ft_insitu, join_by(Pop == Site_ID))

pop_per_env
```

```{r}
pop_above_4_per_env <- pop_per_env |>
  filter(pop_above_4 == T) |>
  drop_na(Type, First_FT_julianday)

pop_above_4_per_env
```

### Statistical model

```{r}
mod_first <- glm(First_FT_julianday ~ Type, data=pop_above_4_per_env, family=gaussian)
summary(mod_first)
car::Anova(mod_first, type="III")
qqnorm(resid(mod_first))
```

```{r}
library(emmeans)  # Für emmGrid-Objekte
library(multcomp) # Für die cld()-Funktion

# function to convert numbers into letter for Tukey comparison of means
convert_to_letters <- function(group) {
  # Remove any spaces in the group string
  group <- gsub(" ", "", group)
  
  # Split the group into individual numbers
  group_numbers <- strsplit(as.character(group), "")[[1]]
  
  # Convert numbers to letters (1 -> a, 2 -> b, ..., 26 -> z)
  group_letters <- sapply(group_numbers, function(x) letters[as.numeric(x)])
  
  # Combine the letters back into a single string
  paste(group_letters, collapse = "")
}

# Calculate least square means and Tukey comparisons
leastsquare <- lsmeans::lsmeans(mod_first, pairwise ~ Type, adjust = "tukey")
tukey_results <- leastsquare$contrasts

# Generate compact letter display
cld_results <- cld(leastsquare)
cld_results

# Extract the compact letter display results
tukey_letters_first <- as.data.frame(cld_results)

# Ensure the Type column is a factor and matches the levels in your data
tukey_letters_first$Type <- factor(tukey_letters_first$Type, levels = levels(pop_above_4_per_env$Type))

#convert numbers to letters
tukey_letters_first$group_letters <- sapply(tukey_letters_first$.group, convert_to_letters)
```

```{r}
ggplot(data = pop_above_4_per_env, mapping = aes(x = Type, y = First_FT_julianday, fill = Type)) +
  geom_boxplot() +
  scale_fill_manual(values = c("meadow" = "#6a176e", "pavement" = "#bc3754", "tree bed" = "#f37819", "wall" = "#f8cd37")) +
  geom_jitter(shape = 21, size = 1.5, fill = "black", position = position_jitter(0.2)) +
  labs(y = "Flowering time (days) in situ", x = NULL) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 14, family = "Times New Roman"),
    axis.title.y = element_text(size = 14, family = "Times New Roman")
  )  +
  geom_text(data = tukey_letters_first, aes(label = group_letters, y = max(pop_above_4_per_env$First_FT_julianday, na.rm = TRUE) + -100), size = 6, family = "Times New Roman")
```

How many replicates do we have per type of environment.

```{r}
pop_per_env_sum <- pop_per_env |>
  filter(pop_above_4 == T) |>
  group_by(Type) |>
  summarise(n_pop = n())
  
pop_per_env_sum
```

------------------------------------------------------------------------

In the data set **FT_table_final** it is Gaëlle data set of the FT CONTROLLED CONDITIONS - ex-situ. She was working with the generation F4 sampled in 2023 and **you are working with the generation F5** from 2025

\- **Column A - H** are the same as your data set!

To extract the FT of the date set you need to do **Sowing date - Flowering date**

Gaëlle also compare the FT in-situ and ex-situ this is the F3/F4 correlation at the end of her script.

```{r}
f4_gaelle <- read_csv(file = "data/FT_table_final.csv")

f4_gaelle
```

```{r}
# A-H are same so reproducing ID from tplant to join
colnames(f4_gaelle) <- paste0(colnames(f4_gaelle), "_F4")
f4_gaelle_w_id <- f4_gaelle |>
  mutate(ID = paste0(as.character(Population_F4), "-", as.character(Replicate_F4)))



tplant_w_f4 <- tplant |>
  left_join(f4_gaelle_w_id, by = "ID")

tplant_w_f4
```

```{r}
ft_insitu
```

```{r}

```
