---
title: "F5 leaf morphology and shoot axis by environment"
format:
  html:
    embed-resources: true
    toc: true
editor: visual
---

```{r}
#| message: false
library(tidyverse)
library(DHARMa)
library(here)
source(here("R/helpers.R"))
set.seed(123)
```

```{r}
#| message: false
tplant <- load_tplant_F5()
```

## EDA of leaf shape and leaf edge per environment

### Populations per ecotype

```{r}
pop_per_eco <- tplant |>
  distinct(pop, .keep_all = TRUE) |>
  select(pop, ecotype) |>
  group_by(ecotype) |>
  summarise(n_pop_per_eco = n())

ggplot(pop_per_eco, aes(y = n_pop_per_eco, x = ecotype, fill = ecotype)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = ecotype_colors) +
  theme_thesis() +
  theme(legend.position = "none") +
  labs(y = "Number of Populations per Type")
```

### Replicates per ecotype

```{r}
repl_per_eco <- tplant |>
  select(replicate, ecotype) |>
  group_by(ecotype) |>
  summarise(n_repl_per_eco = n())

ggplot(repl_per_eco, aes(y = n_repl_per_eco, x = ecotype, fill = ecotype)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = ecotype_colors) +
  theme_thesis() +
  theme(legend.position = "none") +
  labs(y = "Total Individuals of the Populations by Type")
```

### Leaf shape proportions per ecotype

```{r}
leaf_shape_proportions <- tplant |>
  drop_na(leaf_shape) |>
  count(ecotype, leaf_shape, name = "count") |>
  group_by(ecotype) |>
  mutate(proportion = round(count / sum(count), 2)) |>
  select(-count) |>
  pivot_wider(
    names_from   = leaf_shape,
    values_from  = proportion,
    names_prefix = "prop_",
    values_fill  = 0
  ) |>
  ungroup()

leaf_shape_long <- leaf_shape_proportions |>
  select(ecotype, any_of(c("prop_L", "prop_O"))) |>
  pivot_longer(
    cols      = c(prop_L, prop_O),
    names_to  = "leaf_shape",
    values_to = "proportion"
  )

ggplot(leaf_shape_long, aes(x = ecotype, y = proportion, fill = leaf_shape)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  scale_fill_manual(
    values = c("prop_L" = "#6a176e", "prop_O" = "#f8cd37"),
    labels = c("Elongated", "Oval")
  ) +
  labs(x = "Type", y = "Proportion", fill = "Leaf Shape") +
  theme_thesis() +
  theme(legend.position = "top")
```

### Leaf edge proportions per ecotype

```{r}
leaf_edge_proportions <- tplant |>
  drop_na(leaf_edge) |>
  count(ecotype, leaf_edge, name = "count") |>
  group_by(ecotype) |>
  mutate(proportion = round(count / sum(count), 2)) |>
  select(-count) |>
  pivot_wider(
    names_from   = leaf_edge,
    values_from  = proportion,
    names_prefix = "prop_",
    values_fill  = 0
  ) |>
  ungroup()

leaf_edge_long <- leaf_edge_proportions |>
  select(ecotype, any_of(c("prop_G", "prop_S"))) |>
  pivot_longer(
    cols      = c(prop_G, prop_S),
    names_to  = "leaf_edge",
    values_to = "proportion"
  )

ggplot(leaf_edge_long, aes(x = ecotype, y = proportion, fill = leaf_edge)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  scale_fill_manual(
    values = c("prop_G" = "#6a176e", "prop_S" = "#f8cd37"),
    labels = c("Not Serrated", "Serrated")
  ) +
  labs(x = "Type", y = "Proportion", fill = "Leaf Edge") +
  theme_thesis() +
  theme(legend.position = "top")
```

### Leaf edge x shape combinations per ecotype

```{r}
leaf_edge_shape_proportions <- tplant |>
  drop_na(leaf_edge, leaf_shape) |>
  mutate(leaf_edge_shape = paste0(leaf_shape, leaf_edge)) |>
  count(ecotype, leaf_edge_shape, name = "count") |>
  group_by(ecotype) |>
  mutate(proportion = round(count / sum(count), 2)) |>
  select(-count) |>
  pivot_wider(
    names_from   = leaf_edge_shape,
    values_from  = proportion,
    names_prefix = "prop_",
    values_fill  = 0
  ) |>
  ungroup() |>
  select(ecotype, any_of(c("prop_LG", "prop_LS", "prop_OG", "prop_OS")))

leaf_edge_shape_long <- leaf_edge_shape_proportions |>
  pivot_longer(
    cols      = c(prop_LG, prop_LS, prop_OG, prop_OS),
    names_to  = "leaf_edge_shape",
    values_to = "proportion"
  )

ggplot(leaf_edge_shape_long, aes(x = ecotype, y = proportion, fill = leaf_edge_shape)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  scale_fill_manual(
    values = c("prop_LG" = "#440154", "prop_LS" = "#31688e",
               "prop_OG" = "#35b779", "prop_OS" = "#fde725"),
    labels = c("Long Smooth", "Long Serrated", "Oval Smooth", "Oval Serrated")
  ) +
  labs(x = "Type", y = "Proportion", fill = "Leaf Edge Shape") +
  theme_thesis() +
  theme(legend.position = "top")
```

### Mean shoot axis per ecotype

```{r}
axis_per_eco <- tplant |>
  select(n_shoot_axis, ecotype) |>
  drop_na(ecotype, n_shoot_axis) |>
  group_by(ecotype) |>
  summarise(mean_axis = round(mean(n_shoot_axis), 1))

ggplot(axis_per_eco, aes(y = mean_axis, x = ecotype, fill = ecotype)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = ecotype_colors) +
  theme_thesis() +
  theme(legend.position = "none") +
  labs(y = "Average Shoot Axis per Type")
```

### Distribution of sampled populations across ecotypes

```{r}
dist_sampled_pop_per_eco <- tplant |>
  filter(pop_above_3, sampletype != "Not_sampled") |>
  distinct(pop, .keep_all = TRUE) |>
  group_by(ecotype) |>
  summarise(n_sampled_pop = n())

ggplot(dist_sampled_pop_per_eco, aes(y = n_sampled_pop, x = ecotype, fill = ecotype)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = ecotype_colors) +
  theme_thesis() +
  theme(legend.position = "none") +
  labs(y = "Populations with Leafs Sampled from")
```

## MANOVA

Is there a population-level difference in the observed traits across environment types? Because the ~2000 plants are nested in populations, individuals from the same population are non-independent. To address this, a per-population ratio of leaf edge, leaf shape and height is computed.

```{r}
ratio_trait_per_pop <- tplant |>
  drop_na(leaf_shape, leaf_edge) |>
  group_by(pop, ecotype) |>
  summarise(
    n_l                    = sum(leaf_shape == "L"),
    n_o                    = sum(leaf_shape == "O"),
    log_ratio_l_o          = log((n_l + 0.01) / (n_o + 0.01)),
    n_s                    = sum(leaf_edge  == "S"),
    n_g                    = sum(leaf_edge  == "G"),
    log_ratio_s_g          = log((n_s + 0.01) / (n_g + 0.01)),
    above_18cm             = sum(is_above_18cm),
    below_18cm             = sum(!is_above_18cm),
    log_ratio_plant_height = log((above_18cm + 0.01) / (below_18cm + 0.01)),
    .groups = "drop"
  ) |>
  mutate(ecotype = as.factor(ecotype))
```

### MANOVA assumptions

```{r}
table(ratio_trait_per_pop$ecotype)
colnames(ratio_trait_per_pop)

boxplot(log_ratio_l_o          ~ ecotype, data = ratio_trait_per_pop)
boxplot(log_ratio_s_g          ~ ecotype, data = ratio_trait_per_pop)
boxplot(log_ratio_plant_height ~ ecotype, data = ratio_trait_per_pop)

library(mvnormtest)

by(
  ratio_trait_per_pop[, c("log_ratio_l_o", "log_ratio_s_g", "log_ratio_plant_height")],
  ratio_trait_per_pop$ecotype,
  function(x) mshapiro.test(t(x))
)
# Normality violated

library(biotools)
boxM(
  ratio_trait_per_pop[, c("log_ratio_l_o", "log_ratio_s_g", "log_ratio_plant_height")],
  ratio_trait_per_pop$ecotype
)

pairs(ratio_trait_per_pop[, c("log_ratio_l_o", "log_ratio_s_g", "log_ratio_plant_height")])
cor  (ratio_trait_per_pop[, c("log_ratio_l_o", "log_ratio_s_g", "log_ratio_plant_height")])
```

MANOVA assumptions are not met (normality violation). Using PERMANOVA instead.

### PERMANOVA

```{r}
perm_result <- vegan::adonis2(
  ratio_trait_per_pop[, c("log_ratio_l_o", "log_ratio_s_g", "log_ratio_plant_height")] ~ ecotype,
  data = ratio_trait_per_pop,
  method = "euclidean",
  permutations = 999
)

print(perm_result)
```

#### Pairwise PERMANOVA

```{r}
library(pairwiseAdonis) # https://github.com/pmartinezarbizu/pairwiseAdonis

pairwise_result <- pairwise.adonis2(
  ratio_trait_per_pop[, c("log_ratio_l_o", "log_ratio_s_g", "log_ratio_plant_height")] ~ ecotype,
  data = ratio_trait_per_pop,
  method = "euclidean",
  p.adjust.m = "bonferroni"
)
pairwise_result
```

## Leaf shape

Is there a difference in observed leaf shape (Elongated vs Oval) between environment types, controlling for the random effects of tray and population?

### Pop

```{r}
elongated_pop_model <- glm(elongated ~ pop, family = "quasibinomial", data = tplant)
summary(elongated_pop_model)
```

```{r}
anova(elongated_pop_model, test = "F")
```

### Type

```{r}
elongated_type_model <- lme4::glmer(
  elongated ~ ecotype + (1 | pop) + (1 | tray),
  family = binomial(), data = tplant
)
summary(elongated_type_model)
```

```{r}
emmeans::contrast(
  emmeans::emmeans(elongated_type_model, ~ ecotype),
  method = "pairwise", adjust = "tukey"
)
```

```{r}
predictions <- emmeans::emmeans(elongated_type_model, ~ ecotype, type = "response")
cld_result  <- multcomp::cld(predictions, Letters = letters, adjust = "tukey")
```

```{r}
ggplot(cld_result |> as.data.frame(), aes(x = ecotype, y = prob)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.2) +
  geom_text(aes(label = trimws(.group), y = asymp.UCL + 0.1), size = 5) +
  labs(x = "Surface Type", y = "Probability of Elongation") +
  theme_thesis() +
  ylim(0, 1) +
  theme(legend.position = "none")
```

## Leaf edge

Is there a difference in observed leaf edge (Serrated vs Smooth) between environment types, controlling for tray and population?

### Pop

```{r}
serrated_pop_model <- glm(serrated ~ pop, family = "quasibinomial", data = tplant)
summary(serrated_pop_model)
```

```{r}
anova(serrated_pop_model, test = "F")
```

### Type

```{r}
serrated_type_model <- lme4::glmer(
  serrated ~ ecotype + (1 | pop) + (1 | tray),
  family = binomial(), data = tplant
)
summary(serrated_type_model)
```

```{r}
emmeans::contrast(
  emmeans::emmeans(serrated_type_model, ~ ecotype),
  method = "pairwise", adjust = "tukey"
)
```

```{r}
predictions <- emmeans::emmeans(serrated_type_model, ~ ecotype, type = "response")
cld_result  <- multcomp::cld(predictions, Letters = letters, adjust = "tukey")

ggplot(cld_result |> as.data.frame(), aes(x = ecotype, y = prob)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.2) +
  geom_text(aes(label = trimws(.group), y = asymp.UCL + 0.1), size = 5) +
  labs(x = "Surface Type", y = "Probability of Serration") +
  theme_thesis() +
  ylim(0, 1) +
  theme(legend.position = "none")
```

## Shoot axis

```{r}
# hist(tplant$n_shoot_axis)
# left skewed so choosing negative binomial family for glmm
```

```{r}
# axis_model <- glmmTMB::glmmTMB(n_shoot_axis ~ ecotype + (1 | tray) + (1 | pop),
#                                data = tplant,
#                                family = glmmTMB::nbinom2())
#
# summary(axis_model)
```

## Leaf morphology (combined)

```{r}
tplant <- tplant |>
  mutate(leaf_morph = case_when(
    serrated == 1 & elongated == 1 ~ "Serrated & Elongated",
    serrated == 0 & elongated == 1 ~ "Smooth & Elongated",
    serrated == 1 & elongated == 0 ~ "Serrated & Oval",
    serrated == 0 & elongated == 0 ~ "Smooth & Oval",
    TRUE                           ~ NA_character_
  ))
```

```{r}
leaf_type_table <- table(tplant$leaf_morph, tplant$ecotype)
print(leaf_type_table)

chi_test <- chisq.test(leaf_type_table)
print(chi_test)
```

```{r}
leaf_type_table_df <- as.data.frame(leaf_type_table) |>
  rename(ecotype = Var2, morphology = Var1)

ggplot(leaf_type_table_df, aes(x = ecotype, y = Freq, fill = morphology)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c(
    "Serrated & Elongated" = "#6a176e",
    "Serrated & Oval"      = "#f8cd37",
    "Smooth & Elongated"   = "#bc3754",
    "Smooth & Oval"        = "#f37819"
  )) +
  labs(x = "Type", y = "Count of Trait Combination", fill = "Leaf Morphology:") +
  theme_thesis() +
  theme(legend.position = "top")
```

```{r}
corrplot::corrplot(chi_test$residuals, is.cor = FALSE)
```

Red circle means avoided by plant, blue circle means preferred.
