---
title: "F6 leaf morphology by environment"
format:
  html:
    embed-resources: true
    toc: true
editor: visual
---

```{r}
#| message: false
library(tidyverse)
library(here)
source(here("R/helpers.R"))
set.seed(123)
```

```{r}
#| message: false
tplant_F6_germinated <- prepare_f6_germinated()
```

The F6 raw data contains some leaf-edge values beyond the F5 vocabulary:
`WS` (weak serrated) and `H` (half serrated). For the binary serrated /
smooth analyses below they are treated as serrated. The cleaned F6 table
preserved the raw values, so we re-derive `serrated` here from a
simplified edge classification.

```{r}
tplant_F6_germinated <- tplant_F6_germinated |>
  mutate(
    leaf_edge_simple = case_when(
      leaf_edge %in% c("WS", "H") ~ "S",
      TRUE                         ~ leaf_edge
    ),
    serrated = case_when(
      leaf_edge_simple == "S" ~ 1L,
      leaf_edge_simple == "G" ~ 0L,
      TRUE                    ~ NA_integer_
    )
  )
```

## Leaf edge

### Pop

```{r}
serrated_pop2_model <- glm(
  serrated ~ pop, family = "quasibinomial",
  data = tplant_F6_germinated
)
summary(serrated_pop2_model)
```

```{r}
anova(serrated_pop2_model, test = "F")
```

### Type

```{r}
serrated_type2_model <- lme4::glmer(
  serrated ~ ecotype + (1 | tray) + (1 | pop),
  family = binomial(),
  data   = tplant_F6_germinated
)
summary(serrated_type2_model)
```

```{r}
emmeans::contrast(
  emmeans::emmeans(serrated_type2_model, ~ ecotype),
  method = "pairwise", adjust = "tukey"
)
```

```{r}
predictions <- emmeans::emmeans(serrated_type2_model, ~ ecotype, type = "response")
cld_result  <- multcomp::cld(predictions, Letters = letters, adjust = "tukey")

ggplot(cld_result |> as.data.frame(), aes(x = ecotype, y = prob)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.2) +
  geom_text(aes(label = trimws(.group), y = asymp.UCL - 0.1), size = 5) +
  labs(x = "Surface Type", y = "Probability of Serration") +
  theme_thesis() +
  ylim(0, 1) +
  theme(legend.position = "none")
```

## Leaf shape

`elongated` is already encoded in the cleaned F6 table from `leaf_shape ∈ {L, O}`; non-standard values (e.g. `OL`, `O+L`) became NA during cleaning.

### Pop

```{r}
elongated_pop2_model <- glm(
  elongated ~ pop, family = "quasibinomial",
  data = tplant_F6_germinated
)
summary(elongated_pop2_model)
```

```{r}
anova(elongated_pop2_model, test = "F") # genetic component
```

### Type

```{r}
elongated_type2_model <- lme4::glmer(
  elongated ~ ecotype + (1 | tray) + (1 | pop),
  family = binomial(),
  data   = tplant_F6_germinated
)
summary(elongated_type2_model)
```

```{r}
emmeans::contrast(
  emmeans::emmeans(elongated_type2_model, ~ ecotype),
  method = "pairwise", adjust = "tukey"
)
```

```{r}
predictions <- emmeans::emmeans(elongated_type2_model, ~ ecotype, type = "response")
cld_result  <- multcomp::cld(predictions, Letters = letters, adjust = "tukey")

ggplot(cld_result |> as.data.frame(), aes(x = ecotype, y = prob)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.2) +
  geom_text(aes(label = trimws(.group), y = asymp.UCL + 0.1), size = 5) +
  labs(x = "Surface Type", y = "Probability of Elongation") +
  theme_thesis() +
  ylim(0, 1) +
  theme(legend.position = "none")
```

## Shoot axis

```{r}
# hist(tplant_F6_germinated$n_shoot_axis)
# left skewed so choosing negative binomial family for glmm
```

```{r}
# axis2_model <- glmmTMB::glmmTMB(n_shoot_axis ~ ecotype + (1 | tray) + (1 | pop),
#                                 data = tplant_F6_germinated,
#                                 family = glmmTMB::nbinom2())
# summary(axis2_model)
```

## Combined leaf morphology

### Smooth × serrated × elongated × oval

```{r}
tplant_F6_germinated <- tplant_F6_germinated |>
  mutate(leaf_morph = case_when(
    serrated == 1 & elongated == 1 ~ "Serrated & Elongated",
    serrated == 0 & elongated == 1 ~ "Smooth & Elongated",
    serrated == 1 & elongated == 0 ~ "Serrated & Oval",
    serrated == 0 & elongated == 0 ~ "Smooth & Oval",
    TRUE                           ~ NA_character_
  ))
```

```{r}
leaf_type_table2 <- table(tplant_F6_germinated$leaf_morph,
                          tplant_F6_germinated$ecotype)
print(leaf_type_table2)

chi_test2 <- chisq.test(leaf_type_table2)
print(chi_test2)
# Smooth-leaf rows are sparse
```

Not significant — too few plants with smooth leaf margin.

### Excluding the underrepresented smooth class

```{r}
tplant_F6_germinated <- tplant_F6_germinated |>
  mutate(leaf_morph_no_smooth = case_when(
    serrated == 1 & elongated == 1 ~ "Serrated & Elongated",
    serrated == 1 & elongated == 0 ~ "Serrated & Oval",
    TRUE                           ~ NA_character_
  ))
```

```{r}
leaf_type_table3 <- table(tplant_F6_germinated$leaf_morph_no_smooth,
                          tplant_F6_germinated$ecotype)
print(leaf_type_table3)

chi_test3 <- chisq.test(leaf_type_table3)
print(chi_test3)
```

Not significant.

### Proportional bar plot

```{r}
leaf_edge_shape_proportions <- tplant_F6_germinated |>
  mutate(
    serrated_plot  = case_when(serrated  == 1 ~ "S", serrated  == 0 ~ "G"),
    elongated_plot = case_when(elongated == 1 ~ "L", elongated == 0 ~ "O"),
    leaf_edge_shape = paste0(elongated_plot, serrated_plot)
  ) |>
  count(ecotype, leaf_edge_shape, name = "count") |>
  group_by(ecotype) |>
  mutate(proportion = round(count / sum(count), 2)) |>
  select(-count) |>
  pivot_wider(
    names_from   = leaf_edge_shape,
    values_from  = proportion,
    names_prefix = "prop_",
    values_fill  = 0
  ) |>
  ungroup()

leaf_edge_shape_long <- leaf_edge_shape_proportions |>
  select(ecotype, any_of(c("prop_LG", "prop_LS", "prop_OG", "prop_OS"))) |>
  pivot_longer(
    cols      = c(prop_LG, prop_LS, prop_OG, prop_OS),
    names_to  = "leaf_edge_shape",
    values_to = "proportion"
  )

ggplot(leaf_edge_shape_long, aes(x = ecotype, y = proportion, fill = leaf_edge_shape)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  scale_fill_manual(
    values = c("prop_LG" = "#440154", "prop_LS" = "#31688e",
               "prop_OG" = "#35b779", "prop_OS" = "#fde725"),
    labels = c("Long Smooth", "Long Serrated", "Oval Smooth", "Oval Serrated")
  ) +
  labs(x = "Type", y = "Proportion of Leaf Morphology", fill = "") +
  theme_thesis() +
  theme(legend.position = "top")
```

## MANOVA

Is there a population-level difference in observed traits across environment types?

```{r}
ratio_trait_per_pop2 <- tplant_F6_germinated |>
  drop_na(elongated, serrated) |>
  group_by(pop, ecotype, height_stem) |>
  summarise(
    n_l           = sum(elongated == 1),
    n_o           = sum(elongated == 0),
    log_ratio_l_o = log((n_l + 0.01) / (n_o + 0.01)),
    n_s           = sum(serrated  == 1),
    n_g           = sum(serrated  == 0),
    log_ratio_s_g = log((n_s + 0.01) / (n_g + 0.01)),
    .groups = "drop"
  ) |>
  mutate(ecotype = as.factor(ecotype))
```

### MANOVA assumptions

```{r}
table(ratio_trait_per_pop2$ecotype)
colnames(ratio_trait_per_pop2)

boxplot(log_ratio_l_o ~ ecotype, data = ratio_trait_per_pop2)
boxplot(log_ratio_s_g ~ ecotype, data = ratio_trait_per_pop2)
boxplot(height_stem   ~ ecotype, data = ratio_trait_per_pop2)

library(mvnormtest)
by(
  ratio_trait_per_pop2[, c("log_ratio_l_o", "log_ratio_s_g", "height_stem")],
  ratio_trait_per_pop2$ecotype,
  function(x) mshapiro.test(t(x))
)
# Normality violated
```

MANOVA assumptions are not met. Using PERMANOVA instead.

### PERMANOVA

```{r}
perm_result <- vegan::adonis2(
  ratio_trait_per_pop2[, c("log_ratio_l_o", "log_ratio_s_g", "height_stem")] ~ ecotype,
  data = ratio_trait_per_pop2,
  method = "euclidean",
  permutations = 999
)

perm_result
```

Not significantly different.

## Leaf morphology summary

```{r}
tplant_F6_germinated <- tplant_F6_germinated |>
  mutate(leaf_morph = case_when(
    serrated == 1 & elongated == 1 ~ "Serrated & Elongated",
    serrated == 0 & elongated == 1 ~ "Smooth & Elongated",
    serrated == 1 & elongated == 0 ~ "Serrated & Oval",
    serrated == 0 & elongated == 0 ~ "Smooth & Oval",
    TRUE                           ~ NA_character_
  ))

leaf_type_table4 <- table(tplant_F6_germinated$leaf_morph,
                          tplant_F6_germinated$ecotype)
print(leaf_type_table4)
```

```{r}
leaf_type_table4_df <- as.data.frame(leaf_type_table4) |>
  rename(ecotype = Var2, morphology = Var1)

ggplot(leaf_type_table4_df, aes(x = ecotype, y = Freq, fill = morphology)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c(
    "Serrated & Elongated" = "#6a176e",
    "Serrated & Oval"      = "#f8cd37",
    "Smooth & Elongated"   = "#bc3754",
    "Smooth & Oval"        = "#f37819"
  )) +
  labs(x = "Type", y = "Count of Trait Combination", fill = "Leaf Morphology:") +
  theme_thesis() +
  theme(legend.position = "top")
```
